---
title: "Seeking Refuge in Toronto: An Analysis of Refugee Flows in Toronto's Shelter System"
author: "Ruty Korotaev"
date: "February 6, 2023"
abstract: "Amid a growing affordability and housing crisis in Toronto and the Greater Toronto Area (GTA), this paper analyzes the population flow within the city's overnight shelters, using data collected between January 2018 and January 2023. More specifically, this report focuses on understanding and visualizing data related to Toronto's refugee populations, and how members of this demographic are making use of the system. Report findings demonstrate that despite numerous federal and provincial government efforts to  integrate refugees into Canadian society, there is a growing number of refugees that are seeking security and support from Toronto's shelter system."
bibliography: bibliography.bib 
thanks: "Code can be found at https://github.com/rutykorotaev/donaldsonpaper.git"
format: pdf
editor: visual
---

## 1 Introduction

Housing affordability, suburban decline, and gentrification have been key issues plaguing Toronto's housing market over the last 20 years. While this persistent problem is impacting Torontonians across the board, some of Toronto's newest inhabitants, refugees, have been particularly hit hard, leading many to seek support from the city's homeless shelters. Canada has a long and complicated history with refugees, however, since 1980, over one million refugees have settled in Canada and in 2021, Canada became the world's leader in bringing in and welcoming new refugees. Indeed, refugee programming has served as a point of pride for many of Canada's leaders, with Prime Minister Justin Trudeau stating: "Whoever. Wherever. Whenever. Everyone has the right to seek safety." However, while this has been an important priority for the Liberal Party, there are nevertheless many cracks within the resettlement system, which can be seen through the amount of refugees who have sought help and support from Toronto's homeless shelters.

To date, the City of Toronto provides over 7,000 emergency and transitional shelters that are funded by community agencies, with the number of beds steadily growing since 2018. These shelters use the Shelter Management Information System (SMIS) to track population flows and better understand who is making use of these shelters. This system tracks demographic information, such as age, gender, citizenship status, as well as housing data, which detail whether they are actively homeless, have returned from housing or from another shelter, if they are new to the shelter system or returning to it, having previously used the shelter system. While there are countless findings that could be made from this data, this paper will focus solely on the data surrounding refugees and analyzing this population's usage of shelters.

Extant literature about refugees in Toronto has focused largely on health data, including analyses on the mental and physical health of refugees and whether they are seeking support from local services (Dorman et al, 2017; Mule, 2022; Redditt, 2015). Similarly, there has been a great deal of research and scholarly analysis of the housing crisis and gentrification in Toronto, which has led to many people living in overcrowded and poorly maintained accommodations, and others have been altogether priced out of urban housing, being forced to face evictions and homelessness (Wetzstein, 2017). While housing status and an individual's health status can be linked, there is little scholarly information that focuses on how refugees have been seeking support from the shelter system. 

## 2 Data

This paper uses data collected from Open Data Toronto, with to visualize both the demographic information of refugees, as well as their housing situations. As part of this, the main variables that were used in this project include: year, age, and gender, as well as whether the refugees were actively homeless, new to the system or previously known, and whether they have been housed. As with all datasets, there are numerous ethical and statistical issues that may impose a certain bias on the data. This includes the fact that many refugees, and generally those experiencing homelessness, may not be comfortable with providing their details  

This paper uses R (R Core Team 2020) to analyze a dataset from opendatatoronto (Gelfand 2020) which shares data on the population flow within Toronto's shelter system. In addition, several R packages were included in the project, including “tidyverse” (Wickham et al. 2019), “dplyr” (Wickham et al. 2021), “kableExtra” (Zhu 2021) and “knitr” (Xie 2021). 

Figure 1 is a preview of the raw data that was imported through the 'opendatatoronto' library, having extracted the data related just to refugees using homeless shelters.

```{r}
1 + 1
library(tidyverse)
library(knitr)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(knitr)

Refugees_total <- Demographic_Data$gender_female + Demographic_Data$gender_male + Demographic_Data$`gender_transgender,non-binary_or_two_spirit`

dates <- Demographic_Data$`date(mmm-yy)`
percentage_male <- Demographic_Data$gender_male/Demographic_Data$Refugees_total

df <- data.frame(Date = dates, Female = percentage_female, Male = percentage_male, Nonbinary = percentage_nonbinary)

Demographic_Data$date <- as.Date(paste("01-", Demographic_Data1$`date(mmm-yy)`, sep = ""), format = "%d-%b-%y")

Demographic_Data$year <- format(Demographic_Data1$date, format="%y")

#Calculate the annual average of male, female, and non-binary refugees in shelters
male_means <- aggregate(percentage_male ~ year , Demographic_Data, mean)
female_means <- aggregate(percentage_female ~ year , Demographic_Data, mean)
nonbinary_means <- aggregate(percentage_nonbinary ~ year , Demographic_Data, mean)

#Combine all averages into one function and put into chart
cbind(male_means, female_means, nonbinary_means) 
kable(cbind(male_means, female_means, nonbinary_means)
      
```

In addition to the demographic data, this paper also looked the housing status information about the refugees who were staying at homeless shelters. Figures 3 outlines how many refugees were actively homeless and how many were housed. 

```{r}



```


## 4 Discussion

In dissecting and visualizing this dataset, there are numerous insights into the refugee flows within Toronto's shelter system. As it relates to demographic data, it is clear that it is most commonly women refugees who are seeking refuge in the shelter system. Interestingly, as can be seen in Figure 3, many of the refugees, regardless of gender or age, are new to the shelter system and are using it for the first time, which could indicate that many refugees are becoming increasingly vulnerable since resettling in Canada. 

Moreover, when it comes to housing information, there are many findings that can be extracted from the data. Most notably, it appears that a large percentage of refugees in the shelter system are actively homeless, seemingly indicating that government resettlement efforts - which generally guarantee a place to call home - are lacking, and allowing vulnerable people to fall through the cracks. 

Further, while there is no data that indicates the backgrounds and countries of origin for the refugees in shelters, there is a possibility for future literature to delve into how global events and societal upheavals abroad are translating into shifts in demographics throuhgout Toronto's shelter system. For instance, as a more recent example, Russia's escalated invasion of Ukraine has left over 8 million Ukrainians displaced, including many others fleeing the region. As can be seen in Figure 5, there has been a steady growth in refugees in Toronto's shelters since February and March 2022, at the height of the humanitarian crisis in Ukraine when hundreds of thousands of people were fleeing the country. While there is a need to further analyze and gain more demographic information about refugees in Toronto's shelter, it is nevertheless helpful to consider this system within the broader context of world events and its intrinsic linkages with humanitarian efforts throughout the world. 

## References 




