---
title: "Refugees in Toronto Homeless Shelters"
author: "Ruty Korotaev"
abstract: Amid a growing affordability and housing crisis in Toronto, this paper analyzes the flow of people who are using the city's overnight shelters between January 2018 and January 2023. The data demonstrates that ____. 
format: pdf
editor: visual
---

## 1 Introduction

Since \_\_, there has been a growing issue of affordability and housing in Toronto and across the Greater Toronto Area (GTA).

## 2 Data

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## 3 Results

The results of this data set demonstrate that there has been an increase in the amount of refugees in Toronto's shelters, in relation to global events.

## 4 Discussion

In further analyzing the data presented by \_\_\_, it is clear that the flow of refugees in Toronto's shelter system is intrinsically linked to global events.
