---
title: "Seeking Refuge in Toronto: An Analysis of Refugee Flows in Toronto's Shelter System"
author: "Ruty Korotaev"
date: "February 6, 2023"
abstract: "Amid a growing affordability and housing crisis in Toronto and the Greater Toronto Area (GTA), this paper analyzes the population flow within the city's overnight shelters, using data collected by City of Toronto authorities between January 2018 and December 2022. More specifically, this report focuses on understanding how Toronto's refugee population is making use of these shelters, taking into consideration how world events are translating into refugee flows within the local shelter system. Four main variables were explored in this study, including refugee age, gender, housing status, and whether they have been newly identified by shelter operators. The report finds that, following a dip in shelter use during the COVID-19 pandemic, refugees in Toronto are beginning to make use of these services. The report also shows that despite numerous federal and provincial government efforts to integrate refugees into Canadian society, including by providing housing, there is a growing number of refugees that are seeking security and support from Toronto's shelter system."
bibliography: references.bib 
thanks: "Code can be found at https://github.com/rutykorotaev/donaldsonpaper.git"
format: pdf
---

## Introduction

Affordability, inflation, and gentrification have plagued Toronto's housing market over the last 20 years. While these persistent problems are impacting Torontonians across the board, some of Toronto's newest inhabitants, namely refugees, have been hit particularly hard, leading many to seek support from the city's homeless shelters and temporary housing. Canada has a long and complicated history with refugees, and since 1980, the country has resettled over one million refugees. While this has been an important priority for both Canada's federal and provincial governments, there are nevertheless many cracks within the resettlement system, leaving many refugees in precarious situations and in need of secure housing.

To date, the City of Toronto provides over 7,000 emergency and transitional shelters that are funded by community agencies. These shelters use the Shelter Management Information System (SMIS) to track population flows and better understand who is making use of these shelters. This system tracks demographic information, such as age, gender, citizenship status, as well as housing data, which detail whether occupants are actively homeless, have returned from housing or another shelter, and if they are new to the shelter system. While countless findings could be made from this data, this paper will focus solely on the data surrounding refugees and analyzing this population's usage of shelters [@citecityofto].

Extant literature about refugees in Toronto has focused largely on health data, including analyses on the mental [@citemule] and physical [@citedorman] health of refugees. More broadly, there has been a great deal of research and scholarly analysis of the housing crisis and gentrification in Toronto, which has led many people to live in overcrowded and poorly maintained accommodations. Others have been altogether priced out of urban housing, leaving many to face evictions and homelessness [@citeWetzstein].

However, when it comes to understanding how refugees and new immigrants are affected by Toronto's housing crisis, there are only a handful of studies, most of which date back to the early 2000s or 2010s. One article, published in 2008, looked at the experiences of Portuguese-speaking African immigrant groups navigating the city's rental market. Though the data is 15 years old, the study states that most of the refugees interviewed experienced discrimination in their housing search [@citecarlos]. Another report from 2010 argued that, for many refugees, national and local structures in Canada help shape pathways to homelessness. In Toronto specifically, refugee claimants often find themselves in overcrowded and unsafe housing, and are considered to be "particularly vulnerable to homelessness" [@citekissoon]. Overall, while contemporary research on refugees in Toronto may be lacking, presumably many of these issues continue to linger and impact the housing experiences of the city's refugees today.

## Data

This paper uses data collected from Open Data Toronto, and visualizes both the demographic data of refugees, as well as their housing status. As part of this, the main variables that were used in this project include year, age, and gender, as well as whether the refugees were actively homeless and if they are new to the system or previously known. This paper uses R [@citeR] to analyze a dataset from opendatatoronto [@opendatatoronto] which shares data on the population flow within Toronto's shelter system. In addition, several R packages were included in the project, including "tidyverse" [@citetidyverse], "dplyr" [@citedplyr], "kableExtra" [@citekableExtra], and "knitr" [@citeknitr].

To conduct the data cleaning process, we started by loading the original raw dataset and then extracting most data related to the refugee group and the variables we were interested in. We ended up with two datasets that had demographic information, including age and gender, and housing information, which had two variables -- if the refugees were newly identified to the system, and whether they were actively homeless. We also added in columns that included the year sum for each variable, in an effort to visualize the data more clearly, as the raw data included numbers from every month between 2018 to 2022. This process allowed for cleaner visuals that show either the average or sum of each year for any given variable.

@fig-gender depicts the overall numbers of refugees in Toronto's shelters between 2018 and 2022, organized by gender (including those who identified as female, male, and nonbinary).

```{r}
#| echo: false
#| message: false
#| label: fig-gender
#| fig-cap: "Number of Refugees in Toronto's Homeless Shelters by Gender"

library(tidyverse)
library(knitr)

Demographic_Data <- 
  read_csv(here::here("outputs/data/Demographic_Status.csv"))

#Graph data
Demographic_Data |>
  ggplot(aes(x = date, 
             y = number, 
             color = gender)) +
  geom_line() +
  theme_classic() +
  labs(
    x = "Date",
    y = "Number of Refugees"
  )
```

@fig-ages depicts the refugee population by ages, divided into several subgroups.

```{r}
#| echo: false
#| message: false
#| label: fig-ages
#| fig-cap: "Number of Refugees in Toronto's Homeless Shelters by Age"

library(tidyverse)
library(knitr)

Age_Data <- 
  read_csv(here::here("outputs/data/Age_Data.csv"))

#Graph data
Age_Data |>
  ggplot(aes(x = date, 
             y = number, 
             color = ages)) +
  geom_line() +
  geom_point()+
  theme_classic() +
  labs(
    x = "Date",
    y = "Ages of Refugees"
  )

```

@tbl-activehomeless shows the total annual sum of refugees that were reported as being actively homeless when using the shelters between the years 2018 and 2022.

```{r}
#| echo: false
#| message: false
#| label: tbl-activehomeless
#| tbl-cap: "Actively Homeless Refugees Identified in Toronto"

Year_Housing <-
  read_csv(here::here("outputs/data/Year_Housing.csv"))

Year_Housing %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(average_active_homeless = mean(actively_homeless)) %>% 
  knitr::kable(
    col.names = c(
      "Year",
      "Actively Homeless"
    ),
    digits = 0
  )
```

@fig-newlyidentified shows the second part of the housing data section, which looks at how many refugees using homeless shelters were new to the system and had not been previously hosted by Toronto's shelters.

```{r}
#| echo: false
#| message: false
#| label: fig-newlyidentified
#| fig-cap: "Newly Identified Refugees in Toronto"

Housing_Status <- 
  read_csv(here::here("outputs/data/Housing_Status.csv"))

library(ggplot2)

Year_Housing %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(average_newly_identified = mean(newly_identified)) %>% 
  ggplot(aes(x = year, y = average_newly_identified, fill = year)) +
  geom_col() +
  labs(x = "Year", y = "Newly Identified") +
  theme_classic()

```

## Discussion

By visualizing this dataset, there are numerous insights into the refugee flows within Toronto's shelter system. As it relates to demographic data, the number of male and female refugees in shelters, shown in @fig-gender, does not vary greatly. The graph shows that when there is both a decrease or increase of refugees in shelters, the numbers are reflected within both the female and male records. The number of non-binary refugees has remained consistent and relatively low.

Additionally, @fig-ages shows that the 25-44 age group is consistently the most likely to use the shelter system. Notably, the number of refugees in this age group has sky-rocketed throughout 2022, with the numbers being significantly higher than the other age groups. Numbers in the over 65, 16-24, and 44-64 age groups have remained largely consistent, with no large dips as was seen in the 25-44 group.

Moreover, when it comes to housing information, several findings can be extracted from the data. @tbl-activehomeless shows that in 2022, the amount of actively homeless refugees in shelters more than doubled, though it has not reached the pre-pandemic levels. Similarly, @fig-newlyidentified shows that there has been growth in the amount of newly identified refugees in the shelters, which could indicate increased precarity among this demographic given post-pandemic inflation and a rapidly exacerbating housing crisis.

## Analysis

As with all datasets, there are numerous ethical and statistical issues that may impose a certain bias on the data. This includes the fact that many refugees, and generally those experiencing homelessness, may not be comfortable with providing details about their lives and backgrounds with shelter operators, which could lead to skewed samples. Additionally, this data only records the individuals who have sought refuge in Toronto's overnight shelters, and does not include data about other refugees experiencing homelessness or housing insecurity who have not gone to these shelters. Further, while the data is fully anonymized, there may nevertheless be some ethical issues in tracking information on refugees, particularly those who are fleeing oppressive regimes and do not want to be included in public records.

In analyzing this data, there are numerous points that can be highlighted. First, as can be seen in all four charts, the COVID-19 pandemic marked a steep decrease in refugees in shelters. This indicates that shelter authorities either stopped counting inhabitants throughout the pandemic due to staffing shortages and other health precautions, or refugees stopped going to shelters in an effort to avoid getting sick. Nevertheless, in considering this context, it can be argued that while the number of refugees in shelters increased in 2022, it does not necessarily mean that there are more refugees now than there were during the pandemic.

Other global context is also important to consider. While there is no data that indicates the backgrounds and countries of origin of the refugees in shelters, there are several key world events that could be impacting refugee flows. For instance, Russia's escalated invasion of Ukraine in February 2022 has left over 8 million Ukrainians displaced, including many others fleeing the region [@citeukraine]. As can be seen in @fig-newlyidentified, there has been an increase in newly identified refugees in Toronto's shelters since early 2022, which was the height of the humanitarian crisis in Ukraine, when hundreds of thousands of people were fleeing the country. Though there is no way to confirm this without more data on these refugees, future literature could delve into how global events and societal upheavals can impact cities across the world. As such, when considering this data, it is helpful to consider this system within the broader context of world events and the linkages between large- and small-scale humanitarian efforts.

To conclude, it is evident that both the provincial and federal governments need to be doing more to support our society's most vulnerable amid a growing affordability crisis across the country. While the data is not without its problems -- including impacts by the COVID-19 pandemic and lack of representation of Toronto's entire homeless refugee population -- it offers a glimpse into the immense precarity being faced by many in this group and how refugee resettlement programs are allowing thousands of people to fall through the cracks.

## References
