---
title: "Refugees in Toronto Homeless Shelters"
author: "Ruty Korotaev"
abstract: Amid a growing affordability and housing crisis in Toronto, this paper analyzes the flow of people who are using the city's overnight shelters between January 2018 and January 2023. The data demonstrates that
format: pdf
editor: visual
---

## 1 Introduction

Housing affordability, suburban decline, and gentrification have been key issues plaguing Toronto's housing market over the last 20 years. While this persistent problem is impacting Torontonians across the board, some of Toronto's newest inhabitants -- refugees -- have been particularly hit hard, leading many to seek support from the city's homeless shelters. Canada has a long and complicated history with refugees, ranging from the "one is too many" policy that barred Jewish refugees fleeing Nazi Europe to the ---. However, since 1980, over one million refugees have settled in Canada and in 2021, Canada became the world's leader in bringing in and welcoming new refugees. Indeed, refugee programming has served as a point of pride for many of Canada's leaders, with Prime Minister Justin Trudeau stating: "Whoever. Wherever. Whenever. Everyone has the right to seek safety." However, while this has been an important priority for the Liberal Party, there are nevertheless many cracks within the resettlement system, which can be seen through the amount of refugees who have had to seek help and support from Toronto's homeless shelters.

This paper uses RStudio (R Core Team 2020) to analyze a dataset from opendatatoronto (Gelfand 2020) which shares information on the Shelter System Flow within Toronto. As the City of Toronto operates over --- shelters, which offer overnight services, emergency shelters, warming centres, etc., it uses the Shelter Management Information System (SMIS) to better understand who is making use of these shelters, whether they are actively homeless, if they have been previously known to the shelter operators, as well as demographic information, such as their age, gender, and citizenship status. However, this paper will focus solely on the data surrounding refugees and analyzing this population's usage of shelters.

## 2 Data

This paper uses data collected from Open Data Toronto. This paper sought to better understand and explain both the demographic information of refugees, as well as the housing situations of those who were using overnight shelters. Figure 1 depicts how

```{r}
1 + 1
library(tidyverse)
library(knitr)

df <- data.frame(Date = dates, Female=percentage_female, Male = percentage_male, Nonbinary = percentage_nonbinary)
Demographic_Data1$date <- as.Date(paste("01-", Demographic_Data1$`date(mmm-yy)`, sep = ""), format = "%d-%b-%y")
Demographic_Data1$year <- format(Demographic_Data1$date, format="%y")

Demographic_Data2 <- cbind(Demographic_Data1, percentage_male, percentage_female, percentage_nonbinary)
male_means <- aggregate(percentage_male ~ year , Demographic_Data2, mean)
female_means <- aggregate(percentage_female ~ year , Demographic_Data2, mean)
nonbinary_means <- aggregate(percentage_nonbinary ~ year , Demographic_Data2, mean)
cbind(male_means, female_means, nonbinary_means) 
{kable(cbind(male_means, female_means, nonbinary_means))
  2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

In addition, the data demonstrates - refugees by age, refugees housing situation (ie. whether they are actively homeless or rehoused, and how many were new to the system.)

```{r}
##Housing Situation 



```

## 3 Results

Much in line with the initial title of the dataset, the information on refugees demonstrates that there has indeed been a flow of

## 4 Discussion

In further analyzing the data presented by, it is clear that the flow of refugees in Toronto's shelter system is intrinsically linked to global events.
